---
title: "Exploring the BRFSS data"
output: github_document
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data
##About our Data
Our Data is a complete survey of the health status and health issues of adult U.S. citizens which is conducted by BRFSS.The dataset "brfss2013" contains the data of *491775* randomly selected citizens and have a total of *330* variables.
The data is Non - Response bias since the data only comes from people willing to answer the call and going through the long survey

##Process of Data Collection
Data from 50 U.S. states, the District of Columbia, and three U.S. territories is collected and more than 400,000 adult BRFSS surveys are conducted every year. The Survey is taken using landline or celephone telephones

1)  In conducting the BRFSS landline telephone survey, interviewers collect data from a randomly selected adult in a household
2) In conducting the cellular telephone version of the BRFSS questionnaire, interviewers collect data from an adult who participates by using a cellular telephone and resides in a private residence or college housing.

####Scope of inference

##Data is Generalizable
Since the data is collected by sampling random adults from each household. Also none of the researchers interfered with the data hence its an observational study.The data is collected through stratified sampling and the sampling is random hence data is generalizable

##Data is Non Causal
Since the participants were not randomly assigned to groups for specific tasks this cannot be identified as an experimental study and hence no causality can be inferred


* * *

## Part 2: Research questions

**Research quesion 1:**
Is there any correlation between people who smoked at least 100 cigarettes and use smokeless tobacco products Every day, in having other type of cancer(not skin cancer)?
Variables: smoke100,usenow3,chcocncr

Reason for the question: It is believed that people who smoke and use smokeless tobacco are prone to getting cancer, this research question is used to show the statistics in the answers of yes and no with a fair count

**Research quesion 2:**
Is there a correlation between sleeptime, mental health and physical health?
Variables: sleptim1,menthlth,physhlth

Reason for the question: It is believed that people who sleep very less or very much are more prone to having bad physical health and bad mental health, this research question is used to show the statistics of whether there is a correlation or not between the entities

**Research quesion 3:**
Can gender(respondent's sex) and income level be correlated with poor physical or mental health ?
variables: poorhlth,income2,sex

Reason for the question: The question is to brief the possibility that different genders at different income levels experience different levels of healthiness. It is believed that more males than females experience poor health and it is also believed that higher the salary of a person the less is the chance of the person having poor health. These two things can be easily answered with a detailed plot through data analysis.
* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research question 1:**

```{r}
sm_100 <- brfss2013 %>%
  filter(!is.na(smoke100),!is.na(usenow3),!is.na(chcocncr))

##summary stats
cancer100 <- sm_100 %>%
  group_by(chcocncr) %>%
  summarise(smoke_and_tob = sum(smoke100 == 'Yes',usenow3 == 'Every day')) 

cancer100

##plot
ggplot(data = cancer100,aes(x = chcocncr , y = smoke_and_tob ))+geom_bar(stat = 'identity')+ylab('Total Count')+xlab('Other type of cancer and Smoked 100 cigarettes and Use smokeless tobacco every day')
```
##Narative  
As we can see from the given plot and summary stats that citizens of U.S. who smoked 100 cigarettes and use smokeless tobacco everyday got a *11.23%* of chance of getting different types of cancer other than skin cancer which is still a huge amount when generalized to the whole country
We can see in the visuals and stats that *25011* out of *222,649* calculated count had other type of cancer
Hence we can say that the research question proved to be useful and when generalized shows that if you *smoke at least 100 cigarettes* and *use smokeless tobacco products every day* you have a *11.23%* of chance of getting *cancer other than skin cancer*

**Research question 2:**

```{r}
##Creating a small data sample for easy calculations
sampledata = brfss2013[70000:90000,]
sampledata = tbl_df(sampledata)

##Summary Statistics
stats = summarize(sampledata, "sleptim1", mean(sleptim1, na.rm=T))
colnames(stats) = c("Variable", "mean")
physhlthmean = summarize(sampledata, "physhlth", mean(physhlth, na.rm=T))
colnames(physhlthmean) = c("Variable", "mean")
menthlthmean = summarize(sampledata, "menthlth", mean(menthlth, na.rm=T))
colnames(menthlthmean) = c("Variable", "mean")
stats = rbind(stats, physhlthmean,menthlthmean)

stats

##Plot for correlation between sleeptime and physical health
ggplot(sampledata, aes(y= physhlth, x = sleptim1)) + geom_jitter() +geom_smooth()+labs(x="Sleep time in hours / night", y="Bad Physical Health days/month")

##Correlation Value for sleeptime and physical health
cor(sampledata$sleptim1,sampledata$physhlth,use ="complete.obs")


##Plot for correlation between sleeptime and mental health
ggplot(sampledata, aes(y= menthlth, x = sleptim1)) + geom_jitter() +geom_smooth()+labs(x="Sleep time in hours / night", y="Bad Mental Health days/month")

##Correlation value for sleeptime and mental health
cor(sampledata$sleptim1,sampledata$menthlth,use ="complete.obs")
```
##Narative  
In visual representation of physical health and sleeptime and mental health and sleeptime we find that there is no clear correlation or linear relationship and even the correlation value is negative which is not significant, But in the visual representations we can barely see some dips in plotted regression line which shows that people on both extremes can have higher chances of having bad physical or mental health days per month 


**Research question 3:**

```{r}
healthtable <- brfss2013 %>% filter(!(is.na(sex)), !(is.na(poorhlth)), !(is.na(income2)))
healthtable <- healthtable %>% mutate(poorhlth = ifelse(poorhlth > 8, "9+", poorhlth))
healthtable$poorhlth <- factor(healthtable$poorhlth)

##Summary Statistics
healthtable %>% group_by(income2, sex) %>% summarize(count=n())

healthtable %>% group_by(poorhlth, sex) %>% summarize(count=n())

##Plot
plot <- ggplot(healthtable) + aes(x = sex, fill = poorhlth) + geom_bar(position = "fill") + facet_grid(.~income2)
plot <- plot + scale_fill_brewer(name="Poor Health in Last 30 Days", palette = "Spectral") + xlab("Bottom: Respondent's Sex, Top: Income Level") + ylab("Relative Proportion")
plot
```

##Narative  
The research question has been answered,As we can see from the given plot that people with higher salaries have less days of poor health. In the middle segment of the graph of the income level we can clearly see that when it comes to genders, more males experience poor health than women when given the same amount of salary just by a little bit though. We can clearly notice the difference between people who are given high income and people who are given low income. We can also see that females are offered lower salaries in the summary statistics.
